---
title: "Assignment 6"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(tigris)
library(tidycensus)
library(sf)
library(leaflet)
library(mapboxapi)

census_api_key("c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")
```

PART 1 - Analyzing Housing Burden for Renters in Alameda County
```{r}
pums_vars_2018 <- 
  pums_variables %>%
  filter(year == 2018, survey == "acs5")
```

```{r}
Alameda_county <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME == "Alameda")

ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

Alameda_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[Alameda_county, ] %>% 
  st_set_geometry(NULL) %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()

```

```{r, echo =T, results = 'hide'}
ca_pums <- get_pums(
  variables = c(
    "PUMA",
    "GRNTP",
    "SMOCP",
    "ADJHSG",
    "HINCP",
    "ADJINC",
    "TEN"
  ),
  state = "CA",
  year = 2018,
  survey = "acs5"
)

Alameda_pums <-
  ca_pums %>% 
  filter(PUMA %in% Alameda_pumas$PUMACE10)
```


```{r}
burden_threshold <- 0.3

Alameda_burden_renters <-
  Alameda_pums %>% 
  filter(HINCP > 0) %>%
  filter(SPORDER == 1) %>% 
  filter(TEN == 3) %>%
  transmute(
    PUMA = PUMA,
    weight = WGTP,
    housingcost = 
      GRNTP*12*as.numeric(ADJHSG),
    income = HINCP*as.numeric(ADJINC),
    burden_perc = housingcost/income,
    burden_30 = housingcost - burden_threshold*income,
    incomegap_30 = housingcost/burden_threshold - income
  )
```


```{r}
Alameda_burden_renters_pumas <-
  Alameda_burden_renters %>% 
  mutate(
    burdened_30 = ifelse( 
      burden_perc >= burden_threshold,
      weight,
      0
    ),
    excess_30 = ifelse( 
      burden_30 < 0,
      burden_30,
      0
    ),
    burden_30 = ifelse( 
      burden_30 > 0,
      burden_30,
      0
    ),
    incomegap_30 = ifelse( 
      incomegap_30 > 0,
      incomegap_30,
      0
    )
  ) %>% 
  group_by(PUMA) %>% 
  summarize(
    burdened_30 = sum(burdened_30),
    households = sum(weight),
    burden_30 = sum(burden_30*weight),
    incomegap_30 = sum(incomegap_30*weight),
    excess_30 = sum(excess_30*weight)
  ) %>% 
  mutate(
    burdened_30_perc = burdened_30/households
  ) %>% 
  left_join(Alameda_pumas %>% select(PUMA = PUMACE10)) %>% 
  st_as_sf()

sum(Alameda_burden_renters_pumas$burdened_30)/sum(Alameda_burden_renters_pumas$households) #percent of renter households burdened in Alameda County 

sum(Alameda_burden_renters_pumas$burden_30) #total housing burden facing Alameda County renters
```

```{r}
burden_pal1 <- colorNumeric(
  palette = "Purples",
  domain = Alameda_burden_renters_pumas$burdened_30_perc
)

Alameda_burden_renters_pumas %>% 
  leaflet() %>% 
  addMapboxTiles(
    style_id = "streets-v11",
    username = "mapbox"
  ) %>% 
  addPolygons(
    fillColor = ~burden_pal1(burdened_30_perc),
    fillOpacity = 0.5,
    color = "white",
    weight = 0.5,
    label = ~paste0(round(burdened_30_perc*100), "% of households paying 30%+ of income on housing"),
    highlightOptions = highlightOptions(
      weight = 2
    )
  ) %>% 
  addLegend(
    pal = burden_pal1,
    values = ~burdened_30_perc,
    title = "% Cost-burdened<br>households"
  )
```


```{r}
burden_pal2 <- colorNumeric(
  palette = "Reds",
  domain = Alameda_burden_renters_pumas$burden_30/1e6
)

Alameda_burden_renters_pumas %>% 
  leaflet() %>% 
  addMapboxTiles(
    style_id = "streets-v11",
    username = "mapbox"
  ) %>% 
  addPolygons(
    fillColor = ~burden_pal2(burden_30/1e6),
    fillOpacity = 0.5,
    color = "white",
    weight = 0.5,
    label = ~paste0("$", round(burden_30/1e6), "M total annual cost burden"),
    highlightOptions = highlightOptions(
      weight = 2
    )
  ) %>% 
  addLegend(
    pal = burden_pal2,
    values = ~burden_30/1e6,
    title = "Total housing cost<br>burden, in $ millions"
  )
```

Overall, ~49% of renter households in Alameda County are housing burdened (using a burden threshold of 30%). This is almost half of the renter population in the county, showing that housing burden is a serious issue facing many renter households. The map of this housing burden shows that there are a number of PUMAs where more than 49% of renters are housing burdened, and one PUMA in particular shows that 62% of renters are housing burdened. This analysis also shows that ~$1 billion in total annual funding would be needed to eliminate this housing burden for renters in Alameda County. 



PART 2 - Analysis of Parcel Data for one Potrero District Block Group 
```{r}
sf_parcels_shape <- 
  st_read("https://data.sfgov.org/api/geospatial/acdm-wktn?method=export&format=GeoJSON") %>% 
  filter(active == "true") %>% 
  select(
    apn = blklot,
    zoning = zoning_code,
    zoning_desc = zoning_district
  )
```

```{r}
library(readxl)

temp <- tempfile()
download.file("https://sfassessor.org/sites/default/files/uploaded/2020.7.10_SF_ASR_Secured_Roll_Data_2019-2020.xlsx",destfile = temp, mode = "wb")

sf_secured <- read_excel(temp, sheet = "Roll Data 2019-2020")
datakey <- read_excel(temp, sheet = "Data Key")
usecode <- read_excel(temp, sheet = "Class Code Only")

unlink(temp)
```

```{r}
sf_parcels <-
  sf_parcels_shape %>% 
  left_join(
    sf_secured %>% 
      mutate(
        apn = RP1PRCLID %>% 
          str_replace(" ","")
      )
  )
```

```{r}
potrero_bg <-
  block_groups("CA", "San Francisco", cb = T, progress = F) %>%
  filter (GEOID == "060750227022") %>%
  st_transform(4326)

potrero_parcels <- 
  sf_parcels %>% 
  st_centroid() %>% 
  .[potrero_bg, ] %>% 
  st_set_geometry(NULL) %>% 
  left_join(sf_parcels %>% select(apn)) %>% 
  st_as_sf() %>% 
  filter(!is.na(RP1PRCLID))
```

```{r}
potrero_parcels %>%
  leaflet() %>% 
  addMapboxTiles(
    style_id = "streets-v11",
    username = "mapbox"
  ) %>% 
  addPolygons(
    fillColor = "blue",
    color = "black",
    weight = 0.5,
    label = ~zoning
  )
```

```{r}
duplicate_shapes <- 
  potrero_parcels %>% 
  as.data.frame() %>% 
  filter(duplicated(geometry))

condo_parcels <- 
  potrero_parcels %>% 
  filter(geometry %in% duplicate_shapes$geometry)
```

```{r}
potrero_parcels %>%
  st_set_geometry(NULL) %>% 
  group_by(zoning, zoning_desc) %>% 
  summarize(Freq = n())
```


```{r}
potrero_parcels_clean <-
  potrero_parcels %>% 
  filter(zoning != "P") %>% 
  as.data.frame() %>% 
  mutate(geometry = geometry %>% st_as_text()) %>% 
  group_by(geometry) %>% 
  summarize(
    apn = first(apn),
    zoning = first(zoning),
    units = sum(UNITS, na.rm = T),
    stories = max(STOREYNO, na.rm = T),
    floorarea = sum(SQFT, na.rm = T)
  ) %>% 
  ungroup() %>%
  select(-geometry) %>% 
  left_join(potrero_parcels %>% select(apn)) %>% 
  mutate(
    gen_hght = 40
  ) %>%
  st_as_sf()
```


```{r}
projection <- "+proj=utm +zone=10 +ellps=GRS80 +datum=NAD83 +units=ft +no_defs"

potrero_parcels_zoning <-
  potrero_parcels_clean %>% 
  st_centroid() %>% 
  st_set_geometry(NULL) %>% 
  left_join(potrero_parcels_clean %>% select(apn)) %>% 
  st_as_sf() %>% 
  st_transform(projection) %>% 
  mutate(
    lotarea = st_area(.) %>% as.numeric(),
    max_floorarea = case_when(
      zoning %in% c("RH-2","RH-3") ~ lotarea*1.8,
      zoning == "NC-2" ~ lotarea*2.5
    ),
    unused_floorarea = ifelse(
      (max_floorarea - floorarea) > 0,
      (max_floorarea - floorarea),
      0
    ),
    max_units = case_when(
      zoning == "RH-2" ~ 2,
      zoning == "RH-3" ~ 3,
      zoning == "NC-2" ~ lotarea/800
    ),
    unused_units = ifelse(
      (max_units - units) > 0,
      (max_units - units),
      0
    ),
    max_height = ifelse(
      is.na(gen_hght),
      40,
      gen_hght %>% as.numeric()
    ),
    max_stories = floor(max_height/11),
    unused_stories = ifelse(
      (max_stories - stories) > 0,
      (max_stories - stories),
      0
    )
  ) %>% 
  st_transform(4326)
```

```{r}
sum(potrero_parcels_zoning$unused_floorarea, na.rm = T) #total unused floor area

sum(potrero_parcels_zoning$unused_units, na.rm = T) #total number of unused units
```

```{r}
floorarea_pal <- colorBin(
  palette = "Greens",
  bins = c(0,1000,5000,10000,20000,max(potrero_parcels_zoning$unused_floorarea, na.rm = T))
)

leaflet() %>% 
  addMapboxTiles(
    style_id = "light-v9",
    username = "mapbox"
  ) %>% 
  addPolygons(
    data = potrero_parcels_zoning,
    fillColor = ~floorarea_pal(unused_floorarea),
    fillOpacity = 0.75,
    color = "white",
    weight = 0.5,
    label = ~round(unused_floorarea)
  ) %>% 
  addLegend(
    data = potrero_parcels_zoning,
    pal = floorarea_pal,
    values = ~unused_floorarea,
    title = "Unused Floor Area"
  )
```

```{r}
units_pal <- colorBin(
  palette = "Purples",
  bins = c(0,1,2,3,max(potrero_parcels_zoning$unused_units, na.rm = T))
)

leaflet() %>% 
  addMapboxTiles(
    style_id = "light-v9",
    username = "mapbox"
  ) %>% 
  addPolygons(
    data = potrero_parcels_zoning,
    fillColor = ~units_pal(unused_units),
    fillOpacity = 0.75,
    color = "white",
    weight = 0.5,
    label = ~round(unused_units)
  ) %>% 
  addLegend(
    data = potrero_parcels_zoning,
    pal = units_pal,
    values = ~unused_units,
    title = "Additional Units<br>Allowed"
  )
```

There's 764573.9 square feet of floor area that could be built in this area under current zoning, which could create an additional 246 dwelling units. The specific block group of analysis in the Potrero District of San Fransisco is currently only zoned for residential homes (two and three family), small-scale neighborhood commercial, and public use (excluded in this analysis), therefore the extent to which additional dwelling units can be built is limited, but still exists to a small degree. If zoning laws were changed to allow mixed-use housing in this area, the potential for building additional dwelling units would be greater. 
